\section{Licensing}
\label{sect:method:licensing}
One requirement for the parser is to license the produced software under a
permissive open source license. Several such liberal licenses exist, however here we
shall review a short selection of popular and widely adopted licenses, and
choose one for this project. These licenses have been gathered from the Open
Source Initiative (OSI) website\cite{osi_website}, from the category
``\emph{Licenses that are popular and widely used or with strong communities}''.

\subsection*{GPLv2 -- General Public License version 2}
The GPL license is a widely adopted license known for its reciprocality and
enforcement of ``copyleft''\footnote{http://www.gnu.org/copyleft/} to ensure
preservation of software freedom. This license was rejected in favor of more
permissive licenses.

\subsection*{Apache License, Version 2.0}
The Apache 2.0 license is a GPL-compatible\footnote{This is a subject of debate.
The Free Software Foundation argues\cite{fsf_licenses} that the Apache License
2.0 is incompatible  with GPL versions older than version 3 due to patent
clauses in the Apache license} software license proposed by the Apache Software
Foundation. This license is slightly more permissive than the GPL, but was
rejected in favor of a more permissive license. 

\subsection*{New BSD}
BSD-style licenses are named after Berkeley Software Distribution, a Unix-like
operating system. The ``New BSD'' license is a permissive license which
essentially puts the software in the public domain. This is the license chosen
for this project due to its permissive and uncomplicated nature.
