\chapter{Implementation}
\label{chapter:implementation}
This chapter describes the steps made to implement a proof of concept (dubbed
``the prototype'') which showcases some of the most important translation rules
developed in chapter \ref{chapter:translation}. This includes an overall system
description, as well as details about usage of the XQFT Parser. Furthermore, we
describe the process of building a MQL algebra tree, and how the context
sensitive visitor pattern is used. The scoping and symbol table implementation
is covered, as well as how metadata is passed up and down while parsing the
syntax tree and constructing the MQL tree. Finally, this chapter describes in
detail how some of the rules from the ``Tainting Dependencies'' method are
implemented and how they can be made to work in a real-life situation.

\section{Prerequisites}
This proof of concept was implemented in Java 5.0, using regular object
oriented techniques, and is licensed under a liberal BSD license.
This was requested by FAST. Instructions for compilation and installation can
be found in appendix \ref{appendix:installation}.

TODO: den der ``this was requested by'' er litt sner

\section{List Of Supported Features}
This implementation supports the translation of the following XQuery features,
here annotated with references to the descriptions of their respective
translation methods:
\begin{itemize}
  \item FLWOR constructs (section \ref{sect:trans:TD:simpleFLWOR}, page
  \pageref{sect:trans:TD:simpleFLWOR})
  \item Sequence construction (section \ref{sect:trans:TD:seqBuild}, page
  \pageref{sect:trans:TD:seqBuild})
  \item Integer literals (section \ref{sect:trans:TD:literal}, page
  \pageref{sect:trans:TD:litteral})
  \item If-then-else constructs (section \ref{sect:trans:TD:ifThenElse}, page
  \pageref{sect:trans:TD:ifThenElse})
  \item Binary comparison (section \ref{sect:trans:TD:logical}, page
  \pageref{sect:trans:TD:logical})
\end{itemize}

\input{implementation/SystemDesc.tex}
\input{implementation/TheXQFTParser.tex}
\input{implementation/ConstructingMQL.tex}
\input{implementation/ContextSensVisitor.tex}
\input{implementation/ScopingAndSymtab.tex}
\input{implementation/PassingMetadata.tex}
\input{implementation/TaintingDeps.tex}

\section{Summary}
\label{sect:impl:summary}
This chapter has described the implementation of a proof of concept for the
``Tainting Dependencies'' method. In the next chapter, some results are
presented, such as theoretical algebra output. Additionally, algebra generated
by the prototype will be compared to that generated by Pathfinder.
