%THEORY
\chapter{Theory}
\label{chapter:theory}
% Innledning
XML is designed for encapsulation of structured and semistructured data,
relational data, and object repositories. The XQuery query language is designed
to perform flexible queries in such data. A translation of the language into
relational algebra requires detailed knowledge about XQuery for a proper translation to be made. In this chapter, section
\ref{sect:theory:xquery} describes important details about the XQuery
language that are essential to this task. Further, existing implementations of
such translators are documented and compared.

Additionally, the concept of classic relational algebra itself is outlined in
section \ref{sect:theory:relAlg}, as the target algebra will be based on much
of its semantics.

Then in section \ref{sect:parser_and_syntaxtrees}, some common strategies for
parsing and construction of parsers are outlined, as well as techniques for
parsing syntax trees.

Finally, a thoroughly researched method for translating XQuery to relational
algebra dubbed ``Loop Lifting'' is described in section \ref{sect:theory:loop_lifting}.

\input{theory/XQuery.tex}
\input{theory/StateOfTheArt.tex}
\input{theory/RelAlg.tex}
\input{theory/Parsers.tex}
\input{theory/LoopLifting.tex}

\section{Summary}
\label{sect:theory:summary}
%Summary
This chapter has described XQuery and its semantics, including FLWOR
expressions, paths and predicates, and the XQuery Core subset language and its
normalisation rules. Further, existing implementations have been examined and
compared, and we have shown that Pathfinder is one particularly interesting
implementation because of its dependence on a relational back end and thus
relational algebra.

Relational algebra and its semantics were described to set the stage for a
description of the target language which will be presented in the next chapter.

Parser generators and other relevant parser technology has been presented, as
well as some strategies for parsing syntax trees.

Finally, the Loop Lifting method employed by the Pathfinder project to
translate XQuery to relational algebra was described.

In the next chapter, important architectural decisions and methods are
presented.