\renewcommand{\nomname}{Glossary}

\nomenclature{\textbf{DFA}}{\textbf{D}eterministic \textbf{F}inite \textbf{A}utomata. A state machine used typically to formally describe lexical analyzers. For examples see section \ref{sect:antlr:disambiguatingSemantic}.}

\nomenclature{\textbf{AST}}{\textbf{A}bstract \textbf{S}yntax \textbf{T}ree. Two-dimensional trees which encode the structure, as determined by the parser, of the the input symbols. See section \ref{sect:antlr:ast}.}

\nomenclature{\textbf{Ambiguous}}{A grammar is ambiguous if the same sequence of symbols can be interpreted in more than a single way. See section \ref{sect:ambiguousgrammar:ambigTerm}.}

\nomenclature{\textbf{LALR}}{\textbf{L}ook \textbf{A}head \textbf{L}eft to right, \textbf{R}ightmost derivation. CUP (section \ref{sect:method:alternatives}) generates LALR parsers.}

\nomenclature{\textbf{LL}}{\textbf{L}eft to right, \textbf{L}eftmost derivation. ANTLR (chapter \ref{sect:antlr}) generates LL parsers}

\nomenclature{\textbf{W3C}}{\textbf{W}orld \textbf{W}ide \textbf{W}eb \textbf{C}onsortium. An international standards organization for the World Wide Web.}

\nomenclature{\textbf{EBNF}}{\textbf{E}xtended \textbf{B}ackus \textbf{N}aur \textbf{F}orm. A metasyntax notation used to express context-free grammars}

\nomenclature{\textbf{ANTLR}}{\textbf{AN}other \textbf{T}ool for \textbf{L}anguage \textbf{R}ecognition. A predicated-LL(k) parser generator that handles lexers, parsers, and tree parsers.}

\nomenclature{\textbf{Syntactic predicate}}{\emph{ANTLR v3 specific}. A selective form of backtracking used to recognize language constructs that cannot be distinguished with finite look-ahead. See section \ref{sect:antlr:syntacticPredicate}.}

\nomenclature{\textbf{Semantic predicate}}{\emph{ANTLR v3 specific}. A boolean expression used to alter the parse, or throw exceptions based upon runtime information. See section \ref{sect:antlr:semantic_preds}.}

\nomenclature{\textbf{Terminal production}}{A procution that cannot be further divided (in the traditional sense. For ANTLR, see \textbf{fragment}) and generates a token. Also called lexer production. See section \ref{sect:antlr:lexer}}

\nomenclature{\textbf{Fragment}}{\emph{ANTLR v3 specific}. A fragment lexer rule does not represent a complete token, but is a helper rule referenced by another lexer rule. See section \ref{sect:antlr:lexer}.}

\nomenclature{\textbf{Non-terminal production}}{A production that consists of referances to other non-terminals or terminals. Also called parser production. See section \ref{sect:antlr:parser}.}

\nomenclature{\textbf{Production}}{A grammar spesification rule, either terminal or non-terminal.}

\nomenclature{\textbf{Greedy subrule}}{A referenced rule that will match as much input as it can. See \textbf{non-greedy subrule}.}

\nomenclature{\textbf{Non-greedy subrule}}{A referenced rule that will only match input until the refering rule has another alternative that can match the same input. See \textbf{greedy subrule}.}

\nomenclature{\textbf{Hoisting}}{Pulling a predicate out of its enclosing rule and into the prediction expression of another rule. See \textbf{semantic predicate} and sections \ref{sect:antlr:disambiguatingSemantic} and \ref{sect:antlr:gate_semantic_preds}. }

\nomenclature{\textbf{Stdout}}{ Standard output }

\nomenclature{\textbf{Stdin}}{ Standard input }

\nomenclature{\textbf{Stderr}}{ Standard error }

\nomenclature{\textbf{\Large //TODO:}}{dette maa formateres ordentlig}

\nomenclature{\textbf{\Large //TODO:}}{noe mer?}
