\section{FLWOR Expressions}
\label{sect:trans:TD:simpleFLWOR}
The \texttt{let} clause does not explicitly cause any dependencies, only variable binding, and can therefore be
translated into storing the algebraic version of the expression to be bound in the symbol table:
\begin{equation}
\frac{}{\mbox{\texttt{let \$}}\chi \mbox{\texttt{ := }}e \mbox{\texttt{ \ldots}}}\longmapsto
\mbox{\textbf{put(}}\chi\mbox{\textbf{, r(}}e\mbox{\textbf{))}}
\label{rule:trans:TD:letbind}
\end{equation}
The iterator dependences $e.\vartheta$ is stored along with \textbf{r(}$e$\textbf{)} and will piggyback this
algebra tree if it later fetched from the symbol table. If there is more than one variable binding in the
\texttt{let} clause the rule must be applied once per binding as if one binding were one \texttt{let} clause.

As seen by the exerpt of the W3C XQuery EBNF specification in figure \ref{fig:trans:TD:seqEBNF}, a FLWOR
expression may be structured in infinitely many ways. For simplicity and readablilty the translation of such
expressions will be split up in more managable pieces. 

A FLWOR expression may have multiple \texttt{for} clauses, and a \texttt{for} clause may have multiple iterator
variable bindings. This means that one FLWOR may consist of many iterators, the semanics of which is described in
section \ref{sect:method:ast_rewrite}. We assume all possible iterator dependencies generated from the FLWOR, that
is, all iterator variables bound, is stored in a ordered set $\beta$. The dependencies are ordered in the order of
which the corresponding iterator variables are bound, i.e. top down, left to right while parsing the query. When
enclosed in MQL syntax $\beta$ is, as $\vartheta$, to be read as a comma seperated list of the attributes
corresponding to the dependencies.

\begin{figure}[h]
\centering
\begin{tabular}{l}
$[$\texttt{for}/\texttt{let \ldots}$]+$ \\ \quad
$[$\texttt{where }$e_2]?$ \\ \quad
$[$\texttt{order by }$e_3]?$ \\
\texttt{return }$e_4$
\end{tabular}
\label{fig:trans:TD:flworIll}
\caption{Illustration of a FLWOR expression}
\end{figure}

The translation of a single FLWOR like the one illustrated in figure \ref{fig:trans:TD:flworIll} will be executed
as shown in figure \ref{fig:trans:TD:flworExecute}. Firstly, all \texttt{for} and \texttt{let} variables will be
bound as described by the rules \ref{rule:trans:TD:forbind} and \ref{rule:trans:TD:letbind}, respectively.
Further, the \texttt{return} clause will be evaluated. If there is a \texttt{where} clause present, it is
evaluated next, based on the \texttt{where} expression and the result of the \texttt{return} clause, referred to as
\textbf{r(}$e_{C}$\textbf{)}. The order of the items returned from the FLWOR is conditioned by the presens of an
\texttt{order by} clause in the expression. If there is a \texttt{order by} clause, it will order the intermediate
result from the \texttt{return} or \texttt{while} clause, and finalise the FLWOR. If there is no \texttt{order by}
clause, the final evaluation of the FLWOR will be to order the intermediate result according to the iterators.

\begin{figure}[h]
\centering

\begin{tikzpicture}

\node (bind) at (0,0) {$[$\textbf{var binding}$]+$};
\node (return) at (3,0) {\texttt{return} $e_{4}$};
\node (where) at (6,0) {$[$\texttt{where} $e_{2}]?$};
\node (itord) at (9,1.5) {\textbf{iterator order}};
\node (ordby) at (9,-1.5) {$[$\texttt{order by} $e_{3}]?$};
\draw [->,dotted,thick] (bind) to (return);
\draw [->,thick] (return) to (where) node[below,midway] {\tiny\textbf{r(}$e_{C}$\textbf{)}};
\draw [->,thick] (where) to (itord) node[below,midway] {\tiny\textbf{r(}$e_{C}$\textbf{)}};
\draw [->,thick] (where) to (ordby) node[below,midway] {\tiny\textbf{r(}$e_{C}$\textbf{)}};
\end{tikzpicture}
\label{fig:trans:TD:flworExecute}
\caption[FLWOR translation order]{Illustration of step-by-step translation of FLWOR.}
\end{figure}



Any possible iteration, ordering og filtering will be handled by other clauses than the \texttt{return} clause.
The \texttt{return} clause will just ship the evaluated version of the \texttt{return} expression forward:
\begin{equation}
\frac{}{\mbox{\texttt{return }}e_{4}}\longmapsto
\mbox{\textbf{r(}}e_{4}\mbox{\textbf{)}}
\label{rule:trans:TD:returnTaint}
\end{equation}


\subsection{Iterator Ordering}
The \texttt{return} clause is evaluated once for iteration for each of the iterators of the FLWOR expression. The
results of these evaluations are concatenated to form the result of the FLWOR expression. If the \texttt{return}
expression is not dependent on a iterator, its relational form will not have a representation for each
of the iterators iterations and will have to be tainted.

Because each FLWOR iterator creates a new sequence, renumbering is needed. No expression in a sibling or parent
scope of the iterator $I_{\chi}$ may be dependent on $I_{\chi}$. Thus, $\chi$ is not part of the dependencies
$I_{\chi}.\vartheta$ returned from the iterator, and the corresponding $\chi{numb}$ attribute must be removed. 

For each allready bound iterator variable $\chi$, from right to left, bound in an \texttt{for} clause,
the following translation is applied:
\begin{equation}
\frac{}{\mbox{\textbf{iterator order}}}
\longmapsto
\begin{array}{l}
\mbox{\textsf{numberate(index, [}}\beta\mbox{\textsf{, index], [}}\vartheta\mbox{\textsf{];}} \\ \quad
\mbox{\textbf{t(r(}}e_{C}\mbox{\textbf{), }}\beta\mbox{\textbf{)}}
\end{array}
\label{rule:trans:TD:itOrd}
\end{equation}

Where $\vartheta = e_{C}.\vartheta - \chi$.

From equation \ref{eq:trans:TD:taint} it is clear that a expression allready dependent on any iterator in
$\beta$ will not be re-tainted by these iterators.

The \textsf{numberate} operator will have to partition on the remaining dependencies in $\vartheta$ to seperate
the sequences returned from the iterator for all iterators the result is dependent on, as described in section
\ref{sect:trans:TD:implic}.

\begin{myExample}
Consider the query of figure \ref{fig:trans:TD:dblFor}. Notice the expression in the \texttt{return} clause ($e_1$)
is the same as $e_1$ in example \ref{ex:trans:TD:simpleSeq}.

\begin{figure}[h]
\begin{equation*}
\begin{array}{l}
\mbox{\texttt{for \$a in (10,20),}} \\
\mbox{\texttt{\phantom{abcd}\$b in (50,75)}} \\ 
\mbox{\texttt{return }}\underbrace{\mbox{\texttt{(\$a, "no")}} }_{e_{1}}
\end{array}
\end{equation*}
\caption{Example query with two iterators.}
\label{fig:trans:TD:dblFor}
\end{figure}

Expression $e_{1}$ is not depentant on $I_{\mbox{\texttt{b}}}$, and by rule \ref{rule:trans:TD:itOrd} will be
tainted by this iterator. The result of the tainting is shown in figure \ref{fig:trans:TD:final:taint}. The
expression will however not be tainted by $I_{\mbox{\texttt{a}}}$, as it is allready dependent on this iterator
(ref. equation \ref{eq:trans:TD:taint}).

\begin{figure}[h]
\centering
\subfigure[\textbf{t(r(}$e_{1}$\textbf{), \{}\texttt{b}\textbf{\})}]{
\begin{tabular}{|c|c|c|c|} \hline
$bnb$ & $anb$ & $idx$ & $val$ \\ \hline
1 & 1 & 1 & 10 \\ \hline
1 & 2 & 1 & 20 \\ \hline
1 & 1 & 2 & \texttt{"no"} \\ \hline
1 & 2 & 2 & \texttt{"no"} \\ \hline
2 & 1 & 1 & 10 \\ \hline
2 & 2 & 1 & 20 \\ \hline
2 & 1 & 2 & \texttt{"no"} \\ \hline
2 & 2 & 2 & \texttt{"no"} \\ \hline
\end{tabular}
\label{fig:trans:TD:final:taint}
}
\qquad
\subfigure[\textbf{r(}$I_{\mbox{\texttt{a}}}$\textbf{)}]{
\begin{tabular}{|c|c|} \hline
$idx$ & $val$ \\ \hline
1 & 10 \\ \hline
2 & \texttt{"no"} \\ \hline
3 & 10 \\ \hline
4 & \texttt{"no"} \\ \hline
5 & 20 \\ \hline
6 & \texttt{"no"} \\ \hline
7 & 20 \\ \hline
8 & \texttt{"no"} \\ \hline
\end{tabular}
\label{fig:trans:TD:final:rIa}
}
\caption[Example: resolving simple FLWOR]{Evaluating the query in figure \ref{fig:trans:TD:dblFor} yields the
intermediate result (a) and the final result (b). Attribute names are shortened.\label{fig:trans:TD:finalizeExp}}
\end{figure}

With the expression tainted with all possible iterator dependencies of the FLWOR, renumbering is the last
operation required to resolve the query. The \textsf{numberate} operator will sort the tuples first on $anumb$,
then on $bnumb$ and finally $index$ before numbering. As the FLWOR itself does not have any dependencies the
numeration will not be grouped on any attributes. The result of the query is shown in figure
\ref{fig:trans:TD:final:rIa}.
\end{myExample}

\subsection{Where Clause}
The W3C describes the FLWOR expression as generating a tuple stream which contains one tuple for each combination
of values bound in the expression\cite{w3c00}. In this view, the optional \texttt{where} clause serves a filter for
these tuples. The expression in the \texttt{where} clause is evaluated once for tuple. If the boolean value of this
expression is $true$, the tuple is retained, if the boolean value is $false$ the tuple is discarded.

The \texttt{where} clause can be evaluated by only selecting the iteration combinations where the \texttt{where}
expression is true. The result of this will be joined with the result of the \texttt{return} clause. If the result
from the \texttt{return} clause and the \texttt{where} expression has no common iterator dependencies, the
equi-join will have to be replaced by a cartesian product, as discussed in \ref{sect:trans:TD:implic}.

\begin{equation}
\frac{}{\mbox{\texttt{where }}e_2}\longmapsto
\begin{array}{l}
\mbox{\textsf{hhjoin([l.}}(e_{C}.\vartheta \cap e_2.\vartheta)\mbox{\textsf{], [r.}}(e_{2}.\vartheta \cap
e_{C}.\vartheta)\mbox{\textsf{], [l.value, }} \vartheta\mbox{\textsf{];}} \\ \quad
\mbox{\textbf{r(}}e_{C}\mbox{\textbf{)\textsf{;}}} \\ \quad
\mbox{\textsf{select(xqBoolean(value);}} \\ \quad \quad
\mbox{\textbf{r(}}e_{2}\mbox{\textbf{)}\textsf{))}}
\end{array}
\label{rule:trans:TD:where}
\end{equation}

Where $\vartheta = e_{C} \cup e_{2}$, and \textbf{r(}$e_{C}$\textbf{)} is the result of the translation done in
the \texttt{return} clause.

\textsf{l.}$(e_{C}.\vartheta \cap e_2.\vartheta)$ is short for
\textsf{l.}$\chi_1$\textsf{numb,\ldots,l.}$\chi_n$\textsf{numb}, where $e_{C}.\vartheta \cap e_2.\vartheta =
\left\{\chi_1,\ldots,\chi_n\right\}$.

By employing the \textsf{select} operator before the joining of the two relations the number of tuples needed to
be read to execute the join will be minimized. Further, as the \textsf{hhjoin} operator allows for declaring which
attributes to be projected, there is no need for a seperate \textsf{project} operator.

\begin{myExample}
Consider the XQuery query of figure \ref{fig:trans:TD:whereQuery}. The evaluation of $e_{1}$ is very similar to
the evaluation of $e_{1}$ of example \ref{ex:trans:TD:simpleSeq}, and the is shown in figure
\ref{fig:trans:TD:where:re1}. The \texttt{where} clause contains a comparison expression, which will be discussed
later. Figure \ref{fig:trans:TD:where:whereE} shows the result relation of the \texttt{where} expression.

\begin{figure}[h]
\centering
\begin{equation*}
\begin{array}{l}
\mbox{\texttt{for \$a in (10,20)}} \\
\mbox{\texttt{where \$a > 15}} \\
\mbox{\texttt{return}} \underbrace{\mbox{\texttt{(\$a, 14)}}}_{e_{1}}
\end{array}
\end{equation*}
\caption{A FLWOR expression with a \texttt{where} clause \label{fig:trans:TD:whereQuery}}
\end{figure}

\begin{figure}[h]
\centering
\subfigure[\textbf{r(}$e_{1}$\textbf{)}]{
\begin{tabular}{|c|c|c|} \hline
$anb$ & $idx$ & $val$ \\ \hline
1 & 1 & 10 \\ \hline
1 & 2 & 14  \\ \hline
2 & 1 & 20 \\ \hline
2 & 2 & 14  \\ \hline
\end{tabular}
\label{fig:trans:TD:where:re1}
}
\qquad
\subfigure[\textbf{r(}\texttt{\$a > 15}\textbf{)}]{
\begin{tabular}{|c|c|c|} \hline
$anb$ & $idx$ & $val$ \\ \hline
1 & 1 & $false$ \\ \hline
2 & 1 & $true$ \\ \hline
\end{tabular}
\label{fig:trans:TD:where:whereE}
}
\qquad
\subfigure[]{
\begin{tabular}{|c|c|c|} \hline
$anb$ & $idx$ & $val$ \\ \hline
2 & 1 & 20 \\ \hline
2 & 2 & 14  \\ \hline
\end{tabular}
\label{fig:trans:TD:where:intr}
}

\caption[Example: Evaluation of where clause]{Applying rule \ref{rule:trans:TD:where} on (a) and (b)
yields (c). Attribute names are shortened. \label{fig:trans:TD:whereClause}}
\end{figure}

During the evaluation of the FLWOR, the tuples in the \texttt{where} expression relation where $value$ ($val$ in
figure) is not $true$ will be removed. The result is a relation with a single tuple where $anumb$ holds the value
$2$. This result is then joined with the $e_1$ relation on $anumb$, as shown in figure
\ref{fig:trans:TD:where:intr}. Rule \ref{rule:trans:TD:itOrd} will have to be applied to complete the
evaluation of the query.
\end{myExample}


\subsection{Order By Clause Ordering}

As can be seen in figure \ref{fig:trans:TD:ordEBNF}, an \texttt{Order by} clause may contain several specifications on
how to order and what to order by.
\begin{figure}[h]
\begin{Verbatim}
[38] OrderByClause ::= (("order" "by") | ("stable" "order" "by")) 
                           OrderSpecList
[39] OrderSpecList ::= OrderSpec ("," OrderSpec)*
[40] OrderSpec     ::= ExprSingle OrderModifier
[41] OrderModifier ::= ("ascending" | "descending")? 
                           ("empty" ("greatest" | "least"))? 
                           ("collation" URILiteral)?
\end{Verbatim}
\label{fig:trans:TD:ordEBNF}
\caption[W3C EBNF order by clause specification]{W3C EBNF order by clause specification.}
\end{figure}

At this time, Taiting Dependencies only support a simple form of the clause. Only one \texttt{OrderSpec} is
allowed, where the expression to be sorted on will be referred to as $e_4$ -- conformant to figure
\ref{fig:trans:TD:flworExecute}. No \texttt{OrderModifier}s are allowed, and neither is the \texttt{stable}
option. This will be discussed further in sectin \ref{sect:disc:orderby}.

As with the \texttt{where} clause W3C specify the \texttt{order by} clause as operating on a tuple stream created
by the variable bindings in the FLWOR. The clause causes the stream to be reordered into a new, value-based order.
This means that the iterator dependency attributes ($\chi{numb}$) will no longer decide the composition of the
sequence returned from the FLWOR.

Firstly, to ensure that the result will have a representation for all the iterations (still valid after possible
filtering by \texttt{where} clause) of all the expressions iterators, $e_C$ is tainted with $\beta$. The
\texttt{order by} expression is evaluated and joined with the result of the tainting on their common iterator
dependencies. Lastly, this relation is renumbered according to the order of the values stemming from the
\texttt{order by} expression:

\begin{equation}
\frac{}{\mbox{\texttt{order by }}e_3}\longmapsto
\begin{array}{l}
\mbox{\textsf{project(value = l.value, }}\vartheta\mbox{\textsf{;}}\\ \quad
\mbox{\textsf{numberate(index, [r.value, index], [}}\vartheta\mbox{\textsf{];}} \\ \quad \quad
\mbox{\textsf{hhjoin([l.}}(e_{C}.\vartheta \cap e_2.\vartheta)\mbox{\textsf{], [r.}}(e_{2}.\vartheta \cap
e_{C}.\vartheta)\mbox{\textsf{],}} 
\\ \quad \quad \quad \quad \quad\mbox{\textsf{[l.value, r.value, }} \vartheta\mbox{\textsf{];}} \\ \quad \quad
\quad \mbox{\textbf{t(r(}}e_C\mbox{\textbf{), }}\beta\mbox{\textbf{)}\textsf{;}} \\ \quad \quad \quad
\mbox{\textbf{r(}}e_3\mbox{\textbf{)}\textsf{)))}}
\end{array}
\end{equation}

Where $\vartheta = (e_C.\vartheta \cup e_3.\vartheta)-\beta$, and \textbf{r(}$e_{C}$\textbf{)} is the result of
the translation done in the \texttt{where} clause if present, else the \texttt{return} clause.

As desired, all iterator dependency attributes ($\chi{numb}, \chi \in \beta$) will be projected away by the
\textsf{hhjoin} operator. An additional \textsf{project} operator is used to ensure that the $value$ returned from
the expression is the $value$ attribute of the \texttt{return} expression.

\begin{myExample}
Figure \ref{fig:trans:TD:ordQu} shows a query consisting of a FLWOR expression containing a \texttt{order by} clause.
\begin{figure}[h]
\centering
\begin{tabular}{l}
\texttt{for \$a in (4, 6, 8)} \\
\texttt{order by \$a mod 3}\\
\texttt{return \$a}\\
\end{tabular}
\label{fig:trans:TD:ordQu}
\caption{Example query with order by clause}
\end{figure}

The evaluated \texttt{order by} expression is shown in figure \ref{fig:trans:TD:ord:oExp}. How modulus expressions
are translated will be discussed later. After their evaluation, the relational representation of \texttt{order by}
expression relation and the \texttt{return} clause is joined on their common iterator dependencies,
$I_{\mbox{\texttt{a}}}$ ($anumb$). The result of the join is shown in figure \ref{fig:trans:TD:ord:join}. Lastly,
this relation is renumbered sorted on the $r.value$ ($r.val$ in figure) attribute stemming from the \texttt{order by}
expression, as shown in figure \ref{fig:trans:TD:ord:numb}. An additional projection to rename $l.value$ to $value$
is needed to finalise the query.

\begin{figure}[h]
\subfigure[\textbf{r(}\texttt{\$a mod 3}\textbf{)}]{
\begin{tabular}{|c|c|c|} \hline
$idx$ & $anb$ & $val$ \\ \hline
1 & 1 & 1 \\ \hline
1 & 2 & 0 \\ \hline
1 & 3 & 2 \\ \hline
\end{tabular}
\label{fig:trans:TD:ord:oExp}
}
\qquad
\subfigure[]{
\begin{tabular}{|c|c|c|c|} \hline
$l.idx$ & $anb$ & $l.val$ & $r.val$ \\ \hline
1 & 1 & 4 & 1 \\ \hline
1 & 2 & 6 & 0 \\ \hline
1 & 3 & 8 & 2 \\ \hline
\end{tabular}
\label{fig:trans:TD:ord:join}
}
\qquad
\subfigure[]{
\begin{tabular}{|c|c|} \hline
$l.idx$ & $l.val$ \\ \hline
1 & 6 \\ \hline
2 & 4 \\ \hline
3 & 8 \\ \hline
\end{tabular}
\label{fig:trans:TD:ord:numb}
}


\caption{Intermediate results evaluating the query of figure
\ref{fig:trans:TD:ordQu}. (a) shows the evaluated \texttt{order by} expression. (b) shows the \texttt{order by}
expression joined with the \texttt{return} expression. (c) shows the relation in (b) renumbered. Attribute names
are shortened.\label{fig:trans:TD:orderRes}}
\end{figure}


\end{myExample}
