\section{MarkXRemove}
\label{sect:trans:MarkXRemove}

Our original proposition to a method for translating XQuery ASTs into relational algebra was named MarkXRemove.
The rationale behind this name will be explained later. Eventhough it has many shortcomings and flaws, we
will in this section give a quick run through of the method. This is because when we in section
\ref{sect:trans:taintingDependencies} present the Tainting Dependancies method, which is an evolution and a
refinement of MarkXRemove, the new method may be easier understood when seen in the perspective of its origins.
Another reason is that in case of further development of the the Tainted Dependancies method, it may be of help to
also know what will \textit{not} work, what will work partially and \textit{why} it is flawed.

\subsection{Basics}
\label{sect:trans:mxr:basics}

The foundation of the method is that an iterator expression is always translated by calculating the cartesian
product of the iterator sequence and the iterator body, hence the ``X'' in the name. The ``remove'' stems from the
removing of tuples who ended up in the wrong iteration in the cross product result. The cartesian product and the
selection of tuples afterwards actually constitutes a kind of natural join (see section
\ref{sect:theory:relAlg:naturalJoin}) as we will see later.

As the translator comes across an iterator variable declaration, with the variable name $\chi$, it will augment
the representation of the iterator sequence belonging to this variable with an attribute $\chi$\textsf{numb}.
This new attribute will hold consecutive values from 1 to $n$ for a $n$ item long sequence, which will symbolise the
iteration number of the iterator expression seen isolated from possible other surrounding iterator expressions. A
function \texttt{counter()} returning the row number of a relation and a \texttt{project} operator will handle
the augmentation. The corresponding algebra tree is be stored in the symbol table. The ``mark'' of the name of the
method is because of this augmentation.

\subsection{FLWOR}
\label{sect:trans:mxr:flwor}
If the translator later comes accross a reference to an iterator variable $\chi$, it will get the tree from the
symbol table and return it to the referring node without any further ado. The translator is also required to know
which subtrees have a child that has referred to which iterator variable. This is because the $\chi$\textsf{numb}
column could be lost in a \texttt{project} operation without this knowledge.

When the translator returns to the iterator expression node for the variable $\chi$ after traversing the iterator
body, it will, as mentioned before, make sure that the cartesian product between the iterator sequence and body is
calculated. From the result of this, the tuples where the $\chi$\texttt{numb} stemming from the iterator
sequence does not line up with the $\chi$\texttt{numb} stemming from the body are removed. Any tuple with a
$\chi$\texttt{numb} value \texttt{NULL} will be kept.

A \texttt{NULL} value in attribute $\chi$\texttt{numb} for a tuple means that this tuple is not marked, i.e. it is
not dependant on which iteration the $\chi$ iterator expression is in. Unmarked tuples can e.g. stem
from the creation of a sequence. MarkXRemove translates sequence building such as \texttt{(}$e_{1}$\texttt{,
}$e_{1}$\texttt{)} to a simple disjoint union, $r(e_{1})\;\dot\cup\;r(e_{2})$. Where $r()$ symbolises a function
translating XQuery expressions into relational algebra.

The method creates quite simple algebra, as exemplified by the following query:
\begin{Verbatim}
for $i in (1,2,3) return 
  ($i, 'yes')
\end{Verbatim}

which is translated into this algebra:

\begin{Verbatim}
select(ifThenElse(isNull(value), true, eq(r.inumb,l.inumb))
  cross(
    project(inumb = counter(), value;
      make(name:=["value"], [1,2,3]))
    union(
      project(inumb = counter(), value;
        make(name:=["value"], [1,2,3]))
      make(name:=["value"], ['yes']))))
\end{Verbatim} 

\subsection{Flaws}
\label{sect:trans:mxr:flaws}
The main problem with MarkXRemove is its dependancy on particular ordering of tuples in a relation which is a
result of a cartesian product. Not only can not the \texttt{cross} operator promise the particular ordering of its
result, it can not promise any ordering at all. The ordering the method depends on is that for each tuple in the
left relation the tuple is repeated for all tuples in the right relation. This means that any item may apear
anywhere in the resulting sequence, which is not acceptable for evaluation of XQuery expressions where all
sequences are ordered.

Another problem with this method is that any sequence built which includes a reference to an iterator variable is
not fully calculated until the cartesian product between the corresponding relation and the variables iterator
sequence is evaluated. This makes it hard to evaluate expressions where such a sequence is a subexpression. The
iterator \textit{body} of this query:

\begin{Verbatim}
for $i (5,10) return
  ($i, 8) > (6,12)
\end{Verbatim}
would be translated into something like this relational algebra:
\begin{Verbatim}
project(value = gt(l.value, r.value), inumb
cross(
  union(
    project(inumb = counter(), value;
      make(name:=["value"], [5,10]))
    make(name:=["value"], [8]))
  make(name:=["value"], [3,12])))
\end{Verbatim}
which again would produce this relation:

\begin{figure*}[!h]
\centering
\begin{tabular}{|c|c|} \hline
$inumb$ & $value$ \\\hline
1 & \texttt{false} \\\hline
1 & \texttt{false} \\\hline
2 & \texttt{true} \\\hline
2 & \texttt{false} \\\hline
\texttt{NULL} & \texttt{true} \\\hline
\texttt{NULL} & \texttt{false} \\\hline
\end{tabular}
\end{figure*}

The query should of course be evaluated to $(true, true)$, as the \texttt{>} operator in XQuery yields $true$ if
\textit{one} item in the sequence on the left side of the operator is larger than \textit{one} item on the right.
This means that the relation would have to be pruned by a \texttt{select} or \texttt{group}, which can not be done
generally in the relations current state. A possibility would be to postpone the pruning until after the relation
is crossed with the iterator sequence, but there would still not be any not-complex solution. This problem lead to
the introduction of variable dependancy tainting, as we will see in section
\ref{sect:trans:taintingDependancies}.

\underline{\textbf{\LARGE TODO:}} her st\aa r det egentlig noe om numeric predicates, men det er kommentert bort. 
% Within the \texttt{scope} field of a relation returned from a \texttt{lookup} operator lies information about
% which child number the scope is relative to its sibling scopes. E.g. if a \texttt{lookup} had been executed on a
% index holding the following XML-document:
% \begin{Verbatim}
% <a>
%   <b>one</b>
%   <b>two</b>
% </a>
% \end{Verbatim}
% it would be possible to extract that the element containing ``two'' is the second \verb!<b>!-child of its parent.
% And its exactly this information MarkXRemove relies on when translating numeric predicate(see section
% \ref{sect:theory:xquery:Predicates}). But when a numeric predicate is applied to a step expression with a reverse
% axis, e.g. \texttt{ancestor::}, the numbering is reversed. This means that
