\section{Sequence Construction}
\label{sect:trans:TD:seqBuild}

A sequence in XQuery can be built with the comma operator --\texttt{,}. But this operator is the XQuery operator
with the lowest precedence, therefore, in most cases a sequence construction expression will be enclosed in
paratheses. This is to solve parser ambiguities, which can be seen in the
excerpt of the W3C XQuery Full Text EBNF specification\cite{w3c01} in figure \ref{fig:trans:TD:seqEBNF}. An \texttt{ExprSingle} can solely consist of a
\texttt{ParenthesizedExpr} via a series of productions omitted from the figure. Also note a \texttt{ExprSingle}
can be a \texttt{FLWORExpr}.

\begin{figure}[h]
\begin{Verbatim}
[33] FLWORExpr         ::= (ForClause | LetClause)+ WhereClause? 
                               OrderByClause? "return" ExprSingle
[45] IfExpr            ::= "if" "(" Expr ")" "then" ExprSingle 
                               "else" ExprSingle
[31] Expr              ::= ExprSingle ("," ExprSingle)*
[89] ParenthesizedExpr ::= "(" Expr? ")"
\end{Verbatim}
\caption[Excerpt from W3C XQuery EBNF]{Excerpt from W3C XQuery EBNF showing
sequence construction}
\label{fig:trans:TD:seqEBNF}
\end{figure}

As also can be seen from the figure the for-clause of a FLWOR expression, as
many other expressions, accepts an \texttt{ExprSingle}. If a sequence is to be
constructed in the for-clause, it will have to be parenthesised.

With the concept of tainting and iterator dependencies explained, we are now ready to introduce the translation of
an XQuery sequence construction expression:

\begin{equation}
\frac{}{e_{1}\mbox{\texttt{, \ldots, }}e_{n}}\longmapsto
\begin{array}{l}
\mbox{\textsf{numberate(index, [sprIdx, index], [}}\vartheta\mbox{\textsf{];}} \\ \quad
\mbox{\textsf{union(}} \\ \quad \quad
\mbox{\textsf{project(sprIdx=1, index, value; }} \\ \quad \quad \quad
\mbox{\textbf{t(r(}}e_{1}\mbox{\textbf{), }}\vartheta\mbox{\textbf{)}\textsf{);}} \\ \quad \quad
\qquad\vdots\\ \quad \quad
\mbox{\textsf{project(sprIdx=}\textit{n}\textsf{, index, value; }} \\ \quad \quad \quad
\mbox{\textbf{t(r(}}e_{n}\mbox{\textbf{), }}\vartheta\mbox{\textbf{)}\textsf{)))}}
\end{array}
\label{rule:trans:TD:seqConstr}
\end{equation}
Where $\vartheta=e_{1}.\vartheta \cup \ldots \cup e_{n}.\vartheta$.

The basis of a sequence construction is the \textsf{union} operator -- as with MarkXRemove. But because we in
Tainted Dependencies have introduced the explicit ordering of items with the $index$ attribute, additional
operations have been added. Each item in the sequences returned from the subexpressions is equipped with a
temporary field $sprIdx$ (superindex) holding the relative position of each subexpression. Based on the
positioning defined by $sprIdx$ and $index$ the \textsf{numberate} operator can now renumberate the resulting
sequence. The numbering must partition on the fields corresponding to the dependencies in $\vartheta$, to separate
the different sequences constructed for all dependent iterations.

\begin{myExample}
\label{ex:trans:TD:simpleSeq}
\begin{figure}[h]
\begin{equation*}
\begin{array}{l}
\mbox{\texttt{for \$a in (10,20) return}} \\ \quad 
\underbrace{ \mbox{\texttt{(\$a, "no")}} }_{e_{1}}
\end{array}
\end{equation*}
\caption{Simple XQuery query}
\label{fig:trans:TD:simpQuery}
\end{figure}
Consider the simple XQuery query of figure \ref{fig:trans:TD:simpQuery}. Here \textbf{r(}\texttt{\$a}\textbf{)}
will taint \textbf{r(}\texttt{"no"}\textbf{)} with its dependency on the iterator $I_{a}$, the result of which is shown in figure \ref{fig:trans:TD:simpl:ryes}. Further, we can see that
for each iteration of $I_{a}$ the return clause will return a sequence of two items. Having in mind that $anumb$
($anb$ in figure) holds the iteration number of $I_{a}$, this can be seen in figure \ref{fig:trans:TD:simpl:rall}.

\begin{figure}[!h]
\centering
\subfigure[\textbf{r(}\texttt{\$a}\textbf{)}]{
\begin{tabular}{|c|c|c|} \hline
$anb$ & $idx$ & $val$ \\ \hline
1 & 1 & 10 \\ \hline
2 & 1 & 20 \\ \hline
\end{tabular}
\label{fig:trans:TD:simpl:ra}
}
\qquad
\subfigure[\textbf{t(r(}\texttt{"no"}\textbf{),}\textbf{r(}\texttt{\$a}\textbf{)}$.\vartheta$\textbf{)}]{
\begin{tabular}{|c|c|c|} \hline
$anb$ & $idx$ & $val$ \\ \hline
1 & 1 & \texttt{"no"} \\ \hline
2 & 1 & \texttt{"no"} \\ \hline
\end{tabular}
\label{fig:trans:TD:simpl:ryes}
}
\qquad
\subfigure[\textbf{r(}\texttt{(\$a, "no")}\textbf{)}]{
\begin{tabular}{|c|c|c|} \hline
$anb$ & $idx$ & $val$ \\ \hline
1 & 1 & 10 \\ \hline
2 & 1 & 20 \\ \hline
1 & 2 & \texttt{"no"} \\ \hline
2 & 2 & \texttt{"no"} \\ \hline
\end{tabular}
\label{fig:trans:TD:simpl:rall}
}

\caption[Example: constructing a sequence]{Applying translation rule \ref{rule:trans:TD:seqConstr} on (a) and (b)
yields (c). Attribute names are shortened \label{fig:trans:TD:simpleSeq}}
\end{figure}

The sequence construction rule also holds even if the subexpressions of the
soon-to-be sequence are within the body of an iterator the sequence is not
dependent on. Expanding the query of figure \ref{fig:trans:TD:simpQuery} we get
the query of figure \ref{fig:trans:TD:expandQuery}. \begin{figure}[h]
\begin{equation*}
\begin{array}{l}
\mbox{\texttt{for \$a in (10,20) return}} \\ \quad
\mbox{\texttt{for \$b in (50,75) return}} \\ \quad \quad
\underbrace{ \mbox{\texttt{(\$a, "no")}} }_{e_{1}}
\end{array}
\end{equation*}
\caption{Query of figure \ref{fig:trans:TD:simpQuery} expanded}
\label{fig:trans:TD:expandQuery}
\end{figure}
In this query, notice the innermost return clause expression, $e_{1}$, is identical to $e_{1}$ in the previous
query. Here, the result of the sequence construction will still be the relation shown in figure
\ref{fig:trans:TD:simpl:rall}, because $e_{1}.\vartheta=\left\{a\right\}$ -- also as before. $e_{1}$ is not aware
of the iterator $I_{b}$ -- and does not need to be either, as the result of $e_{1}$ is independent of which
iteration number $I_{b}$ is in.

\end{myExample}
