\section{Simplifications}
\label{sect:trans:TD:simplifications}

\textbf{\LARGE TODO:}innledning

In this section we will present some posible simplifications discovered during the develepment of the Tainted
Dependencies method. The $\Rightarrow$ sign is to be read as ``can be written as''. 

\subsection{Literals}
\label{sect:trans:TD:simpl:lit}      

Rule \ref{rule:trans:TD:literal} of section \ref{sect:trans:TD:litteral} shows a very general way to translate
XQuery literals to a relational format. But creating one relation for each litteral is very often unnecessary, and
often quite a bit more resource consuming than alternative solutions. The parent expression of a litteral should
in most cases be informed that its subexpression is a litteral instead of being handed a one-tuple relation.

One such case is if the litteral will be used in a join (predicate-less) or cartesian product. A better solution
will then be to project the litteral into the other relation. Following is an example of how such a expression
should be written:
\begin{equation}
\begin{array}{lcr}
\begin{array}{l}
\mbox{\quad\quad\textsf{\vdots}} \\
\mbox{\textsf{hhjoin([],[],\ldots}} \\ \quad
\mbox{\textsf{\ldots}\textbf{r(}}e\mbox{\textbf{)\ldots}} \\ \quad
\mbox{\textbf{r(}}Literal\mbox{\textbf{)}} \\
\mbox{\quad\quad\textsf{\vdots}}
\end{array}
&
\Rightarrow
&
\begin{array}{l}
\mbox{\quad\quad\textsf{\vdots}} \\
\mbox{\textsf{project(\ldots, rvalue=}}Literal\mbox{\textsf{\ldots}} \\ \quad
\mbox{\textsf{\ldots}\textbf{r(}}e\mbox{\textbf{)\ldots}} \\
\mbox{\quad\quad\textsf{\vdots}}
\end{array}
\end{array}
\label{simpl:trans:TD:joinLit}
\end{equation}

And if the reason why the literal was joined with the relation was because it was part of a arithmetic, comparison
or logical expression, the literal may be moved inside the \textsf{project} operator responsible for executing the
binary operation, shortening the relational algebra even more:

\begin{equation}
\begin{array}{lcr}
\begin{array}{l}
\mbox{\textsf{project(val=OP(l.val, r.val)\ldots}} \\ \quad
\mbox{\textsf{hhjoin([],[],\ldots}} \\ \quad \quad
\mbox{\textsf{\ldots}\textbf{r(}}e\mbox{\textbf{)\ldots}} \\ \quad \quad
\mbox{\textbf{r(}}Literal\mbox{\textbf{)}} \\
\mbox{\quad\quad\textsf{\vdots}}
\end{array}
&
\Rightarrow
&
\begin{array}{l}
\mbox{\textsf{project(val=OP(val, }}Literal\mbox{\textsf{)\ldots}} \\ \quad
\mbox{\textsf{\ldots}\textbf{r(}}e\mbox{\textbf{)\ldots}} \\
\mbox{\quad\quad\textsf{\vdots}}
\end{array}
\end{array}
\label{simpl:trans:TD:binLit}
\end{equation}

If the cases where a literal will have to be translated to a single-tuple relation, in most cases the $index$
attribute will not be needed. But this will probably detected by the optimiser, and may be left out anyway.

\subsection{Sequence Construction}
\label{sect:trans:TD:simpl:seq}

Informing a parent expression if its subexpressions will evaluate to singleton sequences or not can have some
advantages. One is the posibility of detecting certain type errors, as will be discussed in section
\ref{sect:disc:singelton}. Another advandage is gained when it comes to sequence construction.

Rule \ref{rule:trans:TD:seqConstr} in section \ref{sect:trans:TD:seqBuild} describes a general way to build
sequences. But if all the expressions to build a sequence from will evaluate to singleton sequences, there is no
need for the \textsf{numberate} operator. Further, instead of adding $sprIdx$ fields to specify the order, this
can be done directly on the $index$ fields. A version of rule \ref{rule:trans:TD:seqConstr} can therefore be put
like this:
\begin{equation}
\frac{\left\{e_{1},\ldots,e_{n}\right\}\subset{Singletons}}{e_{1}\mbox{\texttt{, \ldots, }}e_{n}}\longmapsto
\begin{array}{l}
\mbox{\textsf{union(}} \\ \quad 
\mbox{\textsf{project(index=1, value; }} \\ \quad \quad 
\mbox{\textbf{t(r(}}e_{1}\mbox{\textbf{), }}\vartheta\mbox{\textbf{)}\textsf{);}} \\ \quad 
\qquad\vdots\\ \quad 
\mbox{\textsf{project(index=}\textit{n}\textsf{, value; }} \\ \quad \quad 
\mbox{\textbf{t(r(}}e_{n}\mbox{\textbf{), }}\vartheta\mbox{\textbf{)}\textsf{))}}
\end{array}
\label{rule:trans:TD:seqIfSingl}
\end{equation}
Where $\vartheta=e_{1}.\vartheta \cup \ldots \cup e_{n}.\vartheta$.

If a sequence construction expression have only literal subexpressions, the translation may be even more
simplified. As the rule stands now, the sequence will be built by splicing together single-tule relations with an
\textsf{union} operator. The \textsf{make} operator does however support multiple items, so a better solution
would be to collect all items in one MQL operator:

\begin{equation}
\frac{\left\{e_{1},\ldots,e_{n}\right\}\subset{Literals}}{e_{1}\mbox{\texttt{, \ldots, }}e_{n}}\longmapsto
\begin{array}{l}
\mbox{\textsf{make(name:=["index", "value"],}} \\ \quad\quad\quad
\mbox{\textsf{[1, \ldots, n], [}}e_{1}\mbox{\textsf{, \ldots, }}e_{n}\mbox{\textsf{])}}
\end{array}
\label{rule:trans:TD:seqIfLit}
\end{equation}

These rules may even be combined, as literals are also singleton sequences. If all the items in the soon to 
be sequence are singletons, all singletons which are literals as well can be inserted into a relation with
the same \textsf{make} operator. The $index$ value of the items will have to be according to their relative 
position within the sequence construction expression. Following is an example of a sequence construction with
only singleton items where not all of them are literals:
\begin{equation*}
\mbox{\texttt{('a', 'b', \$b, 'c')}}\longmapsto
\begin{array}{l}
\mbox{\texttt{union(}} \\ \quad
\mbox{\texttt{project(index=3, value, bnumb;}} \\ \quad \quad
\mbox{\textbf{r(}}\texttt{\$b}\mbox{\textbf{)}\textsf{)}} \\ \quad
\mbox{\textbf{t(}}
\mbox{\textsf{make(name:=["index", "value"],}} \\ \quad \quad \quad \quad \quad 
\mbox{\textsf{[1, 2, 4],['a', 'b', 'c'])}\textbf{,
}}\{\mbox{\texttt{b}}\}\mbox{\textbf{)}\textsf{)}}
\end{array}
\end{equation*}
      

\subsection{FLWORs}
\label{sect:trans:TD:simpl:flwor}        
  FLWOR:
  	\begin{itemize}
        \item  kan bli -> project , ved singelton return
      \end{itemize}

\subsection{Arithmetic Expressions}
\label{sect:trans:TD:simpl:arith}         
  Arithmetic /comp / logical
	\begin{itemize}
      \item multiple +- kan bli dyttet sammen\ldots oddetall - = -, partal = drit i det.
      \item singleton inn i and/or -> slippe select (bull?)
    \end{itemize}
