\section{Existing implementations}
\label{sect:existing_implementations}

Poengtere at disse implementasjonene er fundamentalt annerledes siden vi jobber
mot sm\aa spro indekser og ikke mot XML-traer o.l som er mer vanlig

\textbf{\underline{\LARGE TODO:}} skrive en del mer om tingene, hvordan de jobber p\aa~ ASTen, om de har relasjonell algebra i det heletatt (exist har vel ikke det?)

\subsection{eXist}
eXist\cite{exist_doc} is an open source native XML database with an XQuery
query processor. The eXist system is written in Java. This system stores native
XML data in B-trees and paged files, and document nodes
in persistent DOM trees\cite{exist_factsheet}. Document collections are stored
in a hierarchical manner similar to a regular file system.

eXist has a numerical indexing scheme for identification of relationships
between nodes (parent/child, ancestor/descendant, previous/next sibling). This
provides a structural index for element attribute nodes. In addition, eXist has
a fulltext index for text and attribute values, and range indexes for typed
values.

Based on these provided index types, the eXist XQuery engine relies on path
join algorithms\cite{exist_idx_drv_query} for efficient computation of node
relationships instead of traditional tree traversals. 

\subsection{Pathfinder}
Pathfinder\cite{pathfinderHome} claims to be a ``purely relational XQuery
processor'', which theoretically can utilize any off-the-shelf RBDMS as a
backend for XQuery execution in a relational context.

The technique for transforming XQuery into relational algebra is called ``loop
lifting''\cite{pathfinder_mothertongue}. Loop lifting is a FLWOR-centric
approach which essentially transforms iterations into joins. Loop lifting is
described in detail in section \ref{sect:method:loop_lifting}.

As a relational backend, the XQuery processor uses MonetDB, an integrated
component in the Pathfinder project. However, recent versions of Pathfinder is
also capable of producing SQL code for execution on conventional database
systems. As a proof of concept, they performed the XMark test suite on top of
the Sun DB2 system.

\subsubsection{Loop Lifting}
\begin{itemize}
\item forklare ordentlig hvordan den funker
\end{itemize}

\subsection{Galatex}

\begin{itemize}
\item skriver om til Core
\end{itemize}

\subsection{Common traits}
\begin{itemize}
\item skrive om til Core?
\item relasjonell backend?
\end{itemize}


